### 2023-03-27 response-to-audit-report.md

# DRAFT Response to Audit Report

TODO:

- [ ] Add prefatory text.
- [ ] Add responses and mitigation commits to Isabelle-related findings.
- [ ] Add responses and mitigation commits to Haskell-related findings.
    - [ ] Convert the commit messages to narrative, and just link to the commit.
- [ ] Fix display of equations or replace ones that cannot be fixed with ellipses ("...").
- [ ] Replace broken links and cross references, or use elipses ("...").
- [ ] Verify the section numbering is correct.
- [ ] Add table of contents.
- [ ] Proofread.

## Overview of Isabelle Changes

## Overview of Marlowe-Cardano Specification and Haskell (Plutus) Implementation

TODO: @bwbush will add text here.

```console
commit fca49124f9ebad285bb08d643a3c5fcf5e1df618
Merge: 27a3e32bb 6a0517770
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Wed Mar 15 13:50:51 2023 -0600

    Merge pull request #524 from input-output-hk/PLT-3064

    PLT-3064 Cumulative validator changes based on audit report
```

@@

# 2 Findings

## 2.1 Main concerns

### 2.1.1 Negative deposits allow stealing funds *(Severity: High)*

> **File `marlowe-cardano-specification.md`, `Constraint 6`**
>
> The income from deposits is computed by adding up the deposit inputs, regardless of whether they are negative, while the semantics considers them as zero deposits. Combined with the absence of a balance check on the ending Marlowe state, this allows the ending balance to differ from the value paid to the Marlowe validator.
> 
> This disagreement can be exploited to steal money from a flawed Marlowe contract that allows a negative deposit. The issue is demonstrated in ….

TODO: @bwbush will add text here.

```console
commit 15446073131c4098fed781cdcc701b89b77f6bd0
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Wed Mar 1 16:56:31 2023 -0700

    SCP-5123 Property-based tests for contracts with negative deposits.

commit 4b81ee96b1e7c74c4964cf25f6bb599c80d07a52
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Wed Mar 1 16:56:31 2023 -0700

    SCP-5123 Property-based tests for contracts with negative deposits.

commit 9d5443c5e9306eaf49ac5389df455e07555c9e04
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Thu Mar 2 10:54:48 2023 -0700

    SCP-5123 Improved negative-deposit test case.

commit 20bf9f13fe33acf9612ec26fe640d27975b0dc59
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Wed Mar 1 17:10:11 2023 -0700

    SCP-5123 Rephrased comment.

commit bbb9b8fc21ad93b7bcdb3b60be0f2f9c921f6bec
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Wed Mar 1 16:56:31 2023 -0700

    SCP-5123 Property-based tests for contracts with negative deposits.

commit 597281c176bcb1f326329ba40b0d968d5e7eb84b
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Wed Mar 1 11:36:56 2023 -0700

    SCP-5123 Updated validator script size and hash tests.

commit 68791facc195068717cbc6e55d1e4fdbe1f4a521
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Wed Mar 1 11:20:27 2023 -0700

    SCP-5123 Ensured that negative deposits do not corrupt accounting equation.
```

@@

### 2.1.2 Contracts vulnerable to double satisfaction attacks *(Severity: High)*

> **File `marlowe-cardano-specification.md`, `Constraint 15`**
> 
> No datum is required for outputs fulfilling payments to addresses generated by the evaluation of a Marlowe contract. This implies that these outputs are vulnerable to double satisfaction in transactions involving other contracts that pay to the same wallets. An example is discussed in .
> 
> One way to strengthen the implementation is for the Marlowe validator to demand that outputs paid to addresses contain a datum that identifies the contract instance, like the `TxOutRef` of the validator UTxO being spent. Then cooperation with other contracts is possible without double satisfaction if the validators of the other contracts demand a different datum for their outputs.

TODO: @bwbush will add text here.

```console
commit 0bf75ac775e53507db4e313f705a7a667f98aac1
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Sat Mar 4 09:46:24 2023 -0700

    SCP-5125 Added property-based tests for multiple validators.

commit 9f4fc9cde581b6dbbb939ec9d8e6303b8b67350e
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Fri Mar 3 08:54:31 2023 -0700

    SCP-5141 Property-based tests for duplicate detection in validator.

commit 435ac680eb8119db1c43ec968cf8ea4ef08158b6
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Sat Mar 4 09:46:24 2023 -0700

    SCP-5125 Added property-based tests for multiple validators.

commit 39fcc8aa9d61ed97dfa56e7ce078bf5e4c4cd6cc
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Fri Mar 3 08:54:31 2023 -0700

    SCP-5141 Property-based tests for duplicate detection in validator.

commit 6f6331b463e7f2fddce7ec0f64b37b030f2f795b
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Fri Mar 3 08:54:31 2023 -0700

    SCP-5141 Property-based tests for duplicate detection in validator.

commit 3f222bc363fcf3e9cd2fa0cb5206933ef597a619
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Sat Mar 4 09:46:24 2023 -0700

    SCP-5125 Added property-based tests for multiple validators.

commit 5f673c477ccc9b38626dca505a399962b22c2675
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Sat Mar 4 09:45:32 2023 -0700

    SCP-5125 Included implicit close payments in satisfaction restriction.

commit 4adf115dbf795e4b014eb6ccab0acb20a73e74ed
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Fri Mar 3 13:54:52 2023 -0700

    SCP-5125 Prevented double-satisfaction.
```

@@

### 2.1.3 Missing constructor in equality instance *(Severity: High)*

> **File `Semantics.hs`, Class instance `Eq ReduceWarning`, line *845***
> 
> The constructor `ReduceAssertionFailed` is not mentioned and compares `False` against itself. This might cause validators to fail checking the presence of this particular warning.

TODO: @bwbush will add text here.

```console
commit 04805a393c0815fe0f2a1a6d7f76d9867ffe0c14
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Thu Mar 2 08:27:25 2023 -0700

    SCP-5129 Hardened `Eq` instances in validator files.

    See https://github.com/input-output-hk/marlowe-cardano/pull/513#discussion_r1123062500
    for motivation.

commit 84d65a70c29320fab501ad759bbc71855d1b63fc
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Wed Mar 1 17:21:39 2023 -0700

    SCP-5129 Fixed `(==)` for `ReduceAssertionFailed`.
```

@@

> **File `specification-v3-rc1.pdf`, `Merkleization`**
> 
> There is no property about merkleization, but merkleization is implemented in the Cardano integration.
> 
> Some relevant properties could be:
> 
> 1.  The merkleized contract produces the same payments as the analogous regular contract.
> 2.  If a merkleized case input is applied successfully, it implies that the contract hash in the input corresponds to the continuation of the contract.
> 3.  Merkleizing and unmerkleizing a contract gives back the original contract.

### 2.1.7 Positive balances are not checked for the output state *(Severity: High)*

> **File `marlowe-cardano-specification.md`, `Constraint 13`** 
> 
> *Positive balances* are only checked for the input, not for the output Marlowe state. If the semantics are flawed, a transaction can produce an unspendable output that does not satisfy this constraint.
> 
> If such a transaction is accepted, no further evaluation will be possible since all subsequent transactions will be rejected due to the very same Constraint 13. This is an hypothetical attack vector, where a malicious actor could send a transaction to block a contract.

TODO: @bwbush will add text here.

```console
commit 8855feae473882b82643db792327423152e45b5c
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Thu Mar 2 15:27:42 2023 -0700

    SCP-5141 SCP-5142 Enforced initial and final state validity in validator.

commit ebab31d9ba6dd322d374b34cad3995050a0e476e
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Thu Mar 2 11:52:04 2023 -0700

    SCP-5124 Added property-based test for new final-balance constraint.
```

@@

commit ebab31d9ba6dd322d374b34cad3995050a0e476e
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Thu Mar 2 11:52:04 2023 -0700

    SCP-5124 Added property-based test for new final-balance constraint.
```

### 2.1.9 Total balance of ending state uncomputed *(Severity: High)*

> **File `marlowe-cardano-specification.md`, `Constraint 6`** 
> 
> The constraint says
> 
> > The beginning balance plus the deposits equals the ending balance plus the payments.
> 
> However, the Marlowe validator never computes the total balance of the accounts in the ending Marlowe state. Instead, the ending balance is assumed to be whatever value is paid by the transaction to the Marlowe validator. The natural language should describe precisely what is being checked.

TODO: @bwbush will add text here.

```console
commit 8855feae473882b82643db792327423152e45b5c
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Thu Mar 2 15:27:42 2023 -0700

    SCP-5141 SCP-5142 Enforced initial and final state validity in validator.

commit ebab31d9ba6dd322d374b34cad3995050a0e476e
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Thu Mar 2 11:52:04 2023 -0700

    SCP-5124 Added property-based test for new final-balance constraint.

commit d514bcd075732b07e1c6a73e2e3c68afa01acf2c
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Thu Mar 2 11:52:04 2023 -0700

    SCP-5124 Added property-based test for new final-balance constraint.

commit 7f562545d013dd17472b692b550ffdc7f8a383f3
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Thu Mar 2 11:52:04 2023 -0700

    SCP-5124 Added property-based test for new final-balance constraint.
```

@@

### 2.1.10 Unchecked ending balance *(Severity: High)*

> **File `marlowe-cardano-specification.md`, `Constraint 5`** 
> 
> The balance of the starting Marlowe state is checked to match the value in the input. However, the validator does not check that the ending balance matches the value in the output paid to the Marlowe validator. Similarly to Issue [\[main:positive-balances-unchecked\]](#main:positive-balances-unchecked){reference-type="ref" reference="main:positive-balances-unchecked"}, if there are flaws in the semantics that cause the ending balance to differ from the actual value paid to the validator, this constraint would prevent any transaction from spending the output.
> 
> The specification should at least discuss why the check is absent together with the other similar checks that are not implemented (checking that ending accounts have positive balances, checking that the ending Marlowe state is valid).

```console
commit 0a890845c44ccfe4df97e765e9b9eb743ce7d580
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Thu Mar 2 09:32:09 2023 -0700

    SCP-5124 Ensured script output is consistent with output state.

commit 201c5df9de29b3b142be5ff7d0a9b56d1b378204
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Tue Mar 7 13:35:40 2023 -0700

    PLT-4169 Optimized checking of script output in validator.

commit 26f024e80cea0dd4ebf8bff0f1a10b97aea87894
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Thu Mar 9 12:43:12 2023 -0700

    PLT-3064 Made suggested changes to `examineScripts` in validator.

    See https://github.com/input-output-hk/marlowe-cardano/pull/524#discussion_r1127270744.
```

@@

> **File `Semantics.hs`, Several functions ``** 
> 
> Where `MList.insert` is used in the Isabelle semantics, [AssocMap.insert](https://github.com/input-output-hk/plutus/blob/v1.0.0/plutus-tx/src/PlutusTx/AssocMap.hs#L147-L148) is used in the Cardano implementation. However, the functions are not equivalent as demonstrated by the following examples:
> 
> ``` text
> AssocMap.insert 'a' 1 [('b', 0)] == [('b', 0), ('a', 1)]
> -- whereas
> MList.insert 'a' 1 [('b', 0)] == [('a', 1), ('b', 0)]
> ```
> 
> ``` text
> AssocMap.insert 'b' 1 [('a', 0), (‘b’, 0), (‘b’, 0)] == [('a', 0), ('b', 1), (‘b’, 1)]
> -- whereas
> MList.insert 'b' 1 [(‘a’, 0), ('b', 0), ('b', 0)] == [('a', 0), ('b', 1), ('b', 0)]
> ```
> 
> This renders the Isabelle lemmas inapplicable for the Cardano integration. The lemmas need to demand some properties of an `insert` function without fully spelling it out, or the Cardano integration needs to use `MList.insert` instead of `AssocMap.insert`.
> 
> Similarly, functions `AssocMap.delete` and `MList.delete` differ in behavior when the input map is not sorted:
> 
> ``` text
> AssocMap.delete 'a' [(‘b’, 0), (‘a’, 0)] == [('b', 0)]
> -- whereas
> MList.delete 'a' [(‘b’, 0), ('a', 0)] == [('b', 0), ('a', 0)]
> ```
> 
> Functions `AssocMap.lookup` and `MList.lookup` also differ in behavior when the input map is not sorted:
> 
> ``` text
> AssocMap.lookup 'a' [(‘b’, 0), (‘a’, 0)] == Just 0
> -- whereas
> MList.lookup 'a' [(‘b’, 0), ('a', 0)] == Nothing
> ```
> 
> The following usage places were found:
> -   Line 395, `evalValue` depends on `moneyInAccount` which depends on `AssocMap.lookup`.
> -   Line 413, `evalValue` depends on `AssocMap.lookup`.
> -   Line 428, `evalObservation` depends on `AssocMap.member`.
> -   Line 456, function `updateMoneyInAccount` relies on `AssocMap.delete` and `AssocMap.insert`.
> -   Line 482, function `reduceContractStep` relies on `AssocMap.insert`.
> -   Line 567, function `applyAction` relies on `AssocMap.insert`.

TODO: @bwbush will add text here.

```console
commit 9e068b6a60103f2a40604a7eaea57c5f29b59180
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Wed Mar 1 11:00:45 2023 -0700

    SCP-5126 Aligned `Gen MList` with semantics.

commit b65ccfec3708484ff52514671e0e227c13f084f3
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Wed Mar 1 10:52:17 2023 -0700

    SCP-5126 Hardened equivalence testing for `MList` vs `AssocMap`.

commit a95a616aca59bc7159d3ffa92487f8f021820bff
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Wed Mar 1 10:28:54 2023 -0700

    SCP-5126 Elaborated comments on `MList` property tests.

commit e2277677987f359dcd8754275eccd5b7c9a880a7
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Wed Mar 1 10:23:16 2023 -0700

    SCP-5126 Added note on invariants in unpack and repacking `AssocMap`.

commit af380a292c0369bcc77d8584d81ea80a31326b61
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Tue Feb 28 13:49:44 2023 -0700

    SCP-5126 Property-based tests for `MList` vs `AssocMap`.

commit a2ff6aa334efb7b7955408d29cf0e2edf06bca08
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Tue Feb 28 12:05:32 2023 -0700

    SCP-5126 Annotated Marlowe validator regarding association lists.
```

@@

## 2.5 marlowe-cardano specification

### 2.5.1 Lack of guidelines for creating cooperating contracts *(Severity: Medium)*

> **File `marlowe-cardano-specification.md`, Section `Life Cycle of a Marlowe Contract`** 
> 
> Given that transactions are expected to work with Marlowe and non-Marlowe contracts simultaneously, it would be helpful to offer some guidelines for other contracts to avoid double satisfaction. Some degree of cooperation between the contracts that can appear in the same transaction is unavoidable.
> 
> One measure could be to ask every cooperating contract to refrain from paying to the payout validator. In this way, double satisfaction can not affect the payments of the Marlowe contract, if the Marlowe contract only pays to roles rather than addresses.
> 
> Another alternative would be to demand other contracts' outputs to use datums that are different from the roles used by the Marlowe contract for payments.

TODO: @bwbush will add text here.

```console
commit 00010ea48439892ff9eec54fb93929f56bcb6c7b
Author: Brian W Bush <brian.bush@iohk.io>
Date:   Tue Mar 7 11:20:14 2023 -0700

    PLT-4169 Added text on running Plutus with Marlowe to specification doc.

    This addresses the following audit comment:

    > File marlowe-cardano-specification.md, Section Life Cycle of a
    > Marlowe Contract Given that transactions are expected to work
    > with Marlowe and non-Marlowe contracts simultaneously, it would
    > be helpful to offer some guidelines for other contracts to avoid
    > double satisfaction. Some degree of cooperation between the
    > contracts that can appear in the same transaction is unavoidable.
    >
    > One measure could be to ask every cooperating contract to refrain
    > from paying to the payout validator. In this way, double satisfaction
    > can not affect the payments of the Marlowe contract. Another
    > alternative would be to demand other contracts’ outputs to use datums
    > that are different from the roles used by the Marlowe contract for
    > payments.
```

@@

### 2.5.2 No reference to creating a minting policy *(Severity: Low)*

> **File `marlowe-cardano-specification.md`, Section `Monetary Policy for Role Tokens`** 
> 
> The minting policy is not specified, but a reference needs to be offered to explain how to create one.

TODO: @bwbush will add text here.
